---
title: "The Gravity of Giants"
subtitle: "The 'Magnificent Seven'"
author: "William Orton"
format: 
  revealjs:
    theme: sky
---

```{r}
#| echo: false
library(tidyverse)
library(tidyquant)
```

# What is the S&P 500?

- Stock market index (measures the performance of stocks, like a scoreboard)
- Tracks the largest 500 companies in the United States

# What if **seven companies** in the S&P 500 dominated the [entire]{.underline} market? {style="font-size: 25px; color: #f8fafc; line-height: 1.2;"}

## "The Magnificent Seven"

- Apple
- Amazon
- Microsoft
- Alphabet (owns Google)
- Meta Platforms
- Nvidia
- Tesla

## The Great Market Divide

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5

library(tidyquant)
library(tidyverse)
library(scales)

# Define groups
mag_7 <- c("AAPL", "MSFT", "GOOGL", "AMZN", "NVDA", "META", "TSLA")
spy_ticker <- "SPY"

# Get data
all_data <- tq_get(c(mag_7, spy_ticker), from = "2023-01-01")

# Calculate returns
returns_data <- all_data %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted, mutate_fun = periodReturn, period = "daily", col_rename = "ret") %>%
  mutate(cum_ret = cumprod(1 + ret) - 1)

# Separate mag 7 average from index
mag7_avg <- returns_data %>%
  filter(symbol %in% mag_7) %>%
  group_by(date) %>%
  summarize(cum_ret = mean(cum_ret), group = "Magnificent Seven")

market_bench <- returns_data %>%
  filter(symbol == "SPY") %>%
  select(date, cum_ret) %>%
  mutate(group = "S&P 493 (Rest of Market)")

# plot
ggplot() +
  geom_line(data = mag7_avg, aes(x = date, y = cum_ret, color = group), size = 1.5) +
  geom_line(data = market_bench, aes(x = date, y = cum_ret, color = group), size = 1, linetype = "dashed") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("Magnificent Seven" = "#38bdf8", "S&P 493 (Rest of Market)" = "red")) +
  labs(x = NULL, y = "Cumulative Return", color = NULL) +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```

<div style="font-size: 0.9em;">
- These seven companies are growing way faster than the other 493
- The "Giants" are carrying the weight of the entire market
</div>

## The High Cost of Growth

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# Get prices first
prices <- tq_get(c(mag_7, spy_ticker), from = "2023-01-01")

# Compute daily returns
returns_df <- prices %>%
  group_by(symbol) %>%
  tq_transmute(
    select = adjusted,
    mutate_fun = periodReturn,
    period = "daily",
    col_rename = "ret"
  )

# Rolling 30-day volatility
volatility_df <- returns_df %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(sd = zoo::rollapply(ret, width = 30, FUN = sd, fill = NA, align = "right")) %>%
  ungroup() %>%
  mutate(group = ifelse(symbol == "SPY", "S&P 500", "Mag 7 Average")) %>%
  group_by(group, date) %>%
  summarize(sd = mean(sd, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(volatility_df, aes(x = date, y = sd, color = group)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("#38bdf8", "red")) +
  labs(title = "Rolling 30-Day Standard Deviation of Returns",
       x = NULL, y = "Volatility (SD)", color = NULL) +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```
<div style="font-size: 0.81em;">
- The Mag 7 prices are highly volatile, fluctuating far more than the typical stock
- Rapid growth means investors are in for a much rougher ride
</div>

## Nvidia's Insane Momentum

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# Nvidia golden cross
prices %>%
  filter(symbol == "NVDA") %>%
  filter(date >= "2023-01-01") %>%
  ggplot(aes(x = date, y = adjusted, open = open, high = high, low = low, close = close)) +
  # Using blue for the candlesticks
  geom_candlestick(colour_up = "#38bdf8", colour_down = "red", fill_up = "#38bdf8", fill_down = "red") +
  # 50 day moving average
  geom_ma(ma_fun = SMA, n = 50, color = "#38bdf8", size = 1.25) +
  # 200 day moving average
  geom_ma(ma_fun = SMA, n = 200, color = "black", linetype = "dashed", size = 1) +
  labs(title = "Golden Cross (Nvidia)",
       subtitle = "50-Day SMA (Blue) vs. 200-Day SMA (Dashed Black)",
       x = NULL, y = "Adjusted Price") +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "none"
)
```
<div style="font-size: 0.81em;">
- The aggressive momentum (blue) recently surpassed the long-term average (black)
- This illustrates Nvidia has stepped into "another gear," experiencing extremely rapid growth
</div>

## Pushing Mathematical Limits

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# Bollinger band
prices %>%
  filter(symbol %in% mag_7) %>%
  group_by(date) %>%
  summarize(price = mean(adjusted)) %>%
  # Use tq_mutate to calculate bands manually
  tq_mutate(select = price, 
            mutate_fun = BBands, 
            ma_fun = SMA, sd = 2, n = 20) %>%
  ggplot(aes(x = date, y = price)) +
  # ribbon (shaded area)
  geom_ribbon(aes(ymin = dn, ymax = up), fill = "#38bdf8", alpha = 0.2) +
  # midline (moving average)
  geom_line(aes(y = mavg), color = "#38bdf8", linetype = "dashed", alpha = 0.5) +
  # actual price line
  geom_line(color = "red", size = 1) +
  labs(title = "Bollinger Bands",
       subtitle = "Bollinger Bands on Magnificent Seven Average",
       x = NULL, y = "Price Index") +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```
<div style="font-size: 0.81em;">
- The price line is skirting the upper limit of its "normal" price range
- This means the stock may be extended beyond normal levels
</div>

## Entering the Danger Zone

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# RSi
prices %>%
  filter(symbol %in% mag_7) %>%
  group_by(date) %>%
  summarize(price = mean(adjusted)) %>%
  tq_mutate(select = price, mutate_fun = RSI, n = 14) %>%
  filter(!is.na(rsi)) %>%
  ggplot(aes(x = date, y = rsi)) +
  geom_line(color = "#38bdf8", size = 1) +
  # Overbought & oversold reference lines
  geom_hline(yintercept = 70, linetype = "dashed", color = "#94a3b8") +
  geom_hline(yintercept = 30, linetype = "dashed", color = "#94a3b8") +
  annotate("text", x = as.Date("2023-03-01") +100, y = 75, label = "Overbought (70)", color = "#94a3b8") +
  annotate("text", x = as.Date("2023-03-01") +100, y = 25, label = "Underbought (30)", color = "#94a3b8") +
  labs(title = "RSI",
       subtitle = "14-Day RSI: Magnificent Seven Average",
       x = NULL, y = "RSI Value") +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```
<div style="font-size: 0.81em;">
- An RSI above 70 signals the stock is "red hot" and overbought
- Stocks often crash soon after reaching such high temperatures
</div>

## The Crash

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# drawdown
returns_data %>%
  # use returns data we calculated earlier
  mutate(group = ifelse(symbol %in% mag_7, "Magnificent Seven", "S&P 500")) %>%
  group_by(group, date) %>%
  summarize(ret = mean(ret), .groups = "drop") %>%
  group_by(group) %>%
  mutate(
    cum_ret_index = cumprod(1 + ret),
    running_max   = cummax(cum_ret_index),
    drawdown      = (cum_ret_index / running_max) - 1
  ) %>%
  ggplot(aes(x = date, y = drawdown, fill = group)) +
  geom_area(position = "identity", alpha = 0.5) +
  scale_fill_manual(values = c("Magnificent Seven" = "#38bdf8", "S&P 500" = "red")) +
  scale_y_continuous(labels = percent) +
  labs(title = "Cost of Dominance: Drawdown Analysis",
       subtitle = "Percentage Decline from Previous Peak",
       x = NULL, y = "Drawdown (%)", fill = NULL) +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```
<div style="font-size: 0.81em;">
- This graph reveals how much money investors stand to lose during their "bad" periods
- Although they experience deep drops, these stocks quickly bounce back
</div>

## Correlation Heatmap

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
library(GGally)

# correlation heatmap
all_data %>%
  filter(symbol %in% mag_7) %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted, mutate_fun = periodReturn, period = "monthly", col_rename = "ret") %>%
  pivot_wider(names_from = symbol, values_from = ret) %>%
  select(-date) %>%
  ggcorr(label = TRUE, label_size = 4, label_color = "white", 
         low = "#94a3b8", mid = "#020617", high = "#38bdf8",
         name = "Correlation") +
  labs(title = "Correlation Matrix") +
  theme_minimal(base_size = 16) +
  theme(
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 14),
  legend.text = element_text(size = 14)
)
```
<div style="font-size: 0.81em;">
- These seven stocks all tend to go up and down all at the same time
- One failing company can bring them all crashing down together
</div>

## Monthly Performance Heatmap

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
#monthly return heatmap
returns_data %>%
  mutate(group = ifelse(symbol %in% mag_7, "Magnificent Seven", "S&P 500")) %>%
  group_by(group, month = floor_date(date, "month")) %>%
  summarize(monthly_ret = mean(ret), .groups = "drop") %>%
  ggplot(aes(x = month, y = group, fill = monthly_ret)) +
  geom_tile(color = "#020617", size = 1) +
  scale_fill_gradient2(low = "red", mid = "#94a3b8", high = "#38bdf8", 
                       labels = percent, name = "Return") +
  labs(title = "Monthly Return Clusters: Mag 7 vs. Index",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 22) +
  theme(
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 14),
  legend.text = element_text(size = 14)
)
```
<div style="font-size: 0.81em;">
- This graph indicates which months saw "blue" (good) or "red" (bad) returns
- The patterns show that the growth isn't random but happens in bursts
</div>

## Moving Faster than the Market

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# fixed rolling beta (market sensitivity)
returns_data %>%
  mutate(group = ifelse(symbol %in% mag_7, "Mag7", "SPY")) %>%
  group_by(group, date) %>%
  summarize(ret = mean(ret), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = ret) %>%
  tq_mutate_xy(
    x = Mag7, 
    y = SPY, 
    mutate_fun = rollapply,
    width      = 60,
    FUN        = CAPM.beta,
    by.column  = FALSE,
    col_rename = "rolling_beta"
  ) %>%
  na.omit() %>%
  ggplot(aes(x = date, y = rolling_beta)) +
  geom_line(color = "#38bdf8", size = 1.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "white") +
  annotate("text", x = min(returns_data$date), y = 1.1, 
           label = "More Volatile Than Market", color = "#94a3b8", hjust = 0) +
  labs(title = "60-Day Rolling Beta of Magnificent Seven vs. S&P 500",
       x = NULL, y = "Beta Coefficient") +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20),
  legend.position = "bottom"
)
```
<div style="font-size: 0.81em;">
- These stocks respond more strongly to market news than the S&P 500
- As the market moves an inch, these stocks move a mile
</div>

## High Stakes, High Returns

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
# risk v reward
returns_data %>%
  group_by(symbol) %>%
  tq_performance(Ra = ret, Rb = NULL, performance_fun = table.Stats) %>%
  select(symbol, ArithmeticMean, Stdev) %>%
  mutate(is_mag7 = ifelse(symbol %in% mag_7, "Magnificent Seven", "S&P 500")) %>%
  ggplot(aes(x = Stdev, y = ArithmeticMean, color = is_mag7)) +
  geom_point(aes(size = is_mag7), alpha = 0.7) +
  scale_size_manual(values = c("Magnificent Seven" = 8, "S&P 500" = 3)) +
  scale_color_manual(values = c("Magnificent Seven" = "#38bdf8", "S&P 500" = "red")) +
  theme_minimal(base_size = 24) +
  theme(
  axis.title = element_text(size = 22),
  axis.text = element_text(size = 20),
  legend.text = element_text(size = 20)
)
```
<div style="font-size: 0.81em;">
- You stand to make substantially more money, but you take substantially greater risks of losing money
- The Mag 7 are outliers on the graph, distant from “ordinary” stocks
</div>

# So what?
- Diversification is an illusion right now
- Even if your portfolio seems "diversified," your investments in the S&P 500 almost completely depend on seven technology companies. Whatever happens to these companies will also happen to your investments.

## References & Repos{.smaller}

:::: {.columns}

::: {.column width="60%"} 
- ```bnmonitor``` [vignette](https://github.com/rachwhatsit/BNdiagnostics/blob/master/asia_ex.md)

- Dancho, M., & Vaughan, D. (2026). Package ‘tidyquant’. CRAN. https://CRAN.R-project.org/package=tidyquant

- **Grobys, K. (2025).** *Magnificent 7: unsustainable growth and systemic risk.* Review of Quantitative Finance and Accounting.

-   **Nenkov, D. (2024).** *"The Magnificent Seven" Technology Stocks and Their Impact on the S&P 500: A Review 4 Years Later.* Finance, Accounting and Business Analysis.

:::

::: {.column width="40%"}
#### 👩🏻‍💻Contact

William_Orton1@baylor.edu
Github @williamorton06
:::

::::

## References & Resources {vertical-align="center" .smaller}

:::: {.columns}

::: {.column width="66%"}
### 📚 Academic Sources
* **Grobys, K. (2025).** *Magnificent 7: unsustainable growth and systemic risk.* Review of Quantitative Finance and Accounting.
* **Nenkov, D. (2024).** *"The Magnificent Seven" Technology Stocks and Their Impact on the S&P 500: A Review 4 Years Later.* Finance, Accounting and Business Analysis.

### 📦 Computational Tools
* **tidyquant** [Documentation](https://business-science.github.io/tidyquant/): Integrated financial analysis in the `tidyverse`.
* **ggplot2 & Tidyverse**: For data visualization and exploratory data analysis.
:::

::: {.column width="34%"}
### 👩🏻‍💻 Project Repo
**GitHub** [@williamorton06](https://github.com/williamorton06)

**Contact** William_Orton1@baylor.edu

**Data Source** Daily log-price data via Yahoo Finance.
:::

::::